---
title: "Evaluation of Invasive Plant Species Composition and Structural Diversity on the Big Island of Hawai'i"
author: "Environmental Data Analytics | Elizabeth Miles-Flynn & Irina Roybal"
subtitle: "Mauna Loa, Ola'a, and Nahuku/East Rift"
date: "December 12, 2025"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    toc_float: true
    lof: true
    theme: cerulean
    highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
##Note: These weren't learned in class yet helpful to create list of figures in html - used online sources. 
#install.packages("xfun")
#install.packages("bookdown")

#Setup
library(bookdown)
library(tidyverse)
library(lubridate)
library(here)
library(knitr)
library(sf)
library(trend)
library(zoo)

#Check Working Directory
getwd()
here()

#Set Theme
hawaiitheme <- 
   theme(
    plot.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    legend.key = element_rect(fill = "white", color = NA),
    legend.title = element_text(color = "black", 
                                face = "bold",
                                family = "serif"),
    legend.text = element_text(color = "black",
                               family = "serif"),
    panel.background = element_rect(fill = "cyan3"),
    panel.grid.major = element_line(color = "lightblue2"),
    panel.grid.minor = element_line(color = "lightblue2"),
    plot.title = element_text(face = "bold",
                              size = 16,
                              color = "black",
                              family = "serif"),
    axis.title.x  = element_text(
                               face = "italic", 
                               size = 8, 
                               color = "black",
                               family = "serif"),
    axis.title.y = element_text( 
                               face = "italic", 
                               size = 8, 
                               color = "black",
                               family = "serif"))
```

## Git Repository

<https://github.com/irinaroybal/MilesflynnRoybal_ENV872_EDA_FinalProject.git>

## List of Figures

**Figure** \@ref(fig:Image): Landsat mosaic of the Big Island of Hawaii,
elevation in meters

**Figure** \@ref(fig:MapofSamplingSites): Map showing the sampling sites
are color-coded by region

**Figure** \@ref(fig:ResearchQ1): Overall invasive species richness at
each Site on the Big Island of Hawai’i

**Figure** \@ref(fig:ResearchQ2): Invasive species richness by
vegetation type at each site on the Big Island of Hawai’i

**Figure** \@ref(fig:ResearchQ3): 2010 to 2023 Detection of invasive
species for each site on the Big Island of Hawai’i

**Figure** \@ref(fig:SpecificSpeciesMappingExploration): Locations of
highly invasive species on the Big Island

**Figure** \@ref(fig:fern): Invasive fern species abundance at each site
on the Big Island of Hawai’i

**Figure** \@ref(fig:grass): Invasive grass species abundance at each
site on the Big Island of Hawai’i

**Figure** \@ref(fig:herb): Invasive herb species abundance at each site
on the Big Island of Hawai’i

**Figure** \@ref(fig:rush): Invasive rush species abundance at each site
on the Big Island of Hawai’i

**Figure** \@ref(fig:sedge): Invasive sedge species abundance at each
site on the Big Island of Hawai’i

**Figure** \@ref(fig:shrub): Invasive shrub species abundance at each
site on the Big Island of Hawai’i

**Figure** \@ref(fig:tree): Invasive tree species abundance at each site
on the Big Island of Hawai’i

**Figure** \@ref(fig:treefern): Invasive tree fern species abundance at
each site on the Big Island of Hawai’i

**Figure** \@ref(fig:vine): Invasive vine species abundance at each site
on the Big Island of Hawai’i

*\*For section "4.3 Diversity of Invasive Species Across Habitat Type",
tab numbers do not reflect the numbers of the figures, which are present
in the figure caption underneath each graph.*

# Rationale and Research Questions

The purpose of this study is to analyze the spatial distribution of
invasive species across three study sites on the Big Island of Hawai‘i.
Prior to Polynesian arrival in the remote Pacific, new invasive species
established naturally at an estimated rate of one species every 50,000
years. After Polynesian settlement, this rate increased to roughly three
to four species every hundred years (Young 2016). Following European
contact, the introduction rate accelerated dramatically to approximately
twenty invasive species per year in Hawai‘i (Ewell 1999). While
non-native species include any organisms introduced outside their
natural range, only a subset of them become invasive. Invasive species
are defined as non-native organisms that spread rapidly and/or cause
ecological, economic, or cultural harm. These invasive species pose a
substantial threat to Hawai‘i’s unique native biota. Of the roughly
10,000 native species found across the Hawaiian Islands, nearly 90% are
endemic—occurring nowhere else on Earth (Cordell 2021). Our goal is to
evaluate the invasive species on the Big Island across three study sites
over the past thirteen years to inform best management practices to
protect endemic species of the Hawaiian Islands. Our dataset focuses on
three sites on the Big Island: Mauna Loa, Ola’a, and Nahuku/East Rift.
Mauna Loa is classified as a subalpine shrubland, and Ola’a and
Nahuku/East Rift are considered wet forest habitats. Our dataset ranges
from 2010 to 2023 and contains coordinates for each survey site where a
transect was used to count species-level observations and counts of
repeated occurrences.

**Through our analysis, we aim to answer the following research
questions:**

1.  **Does invasive species' richness change across the three sample
    regions?**

2.  **Does invasive species' structural diversity change across the
    three sample regions?**

3.  **How does the frequency and distribution of invasive species change
    over time for each site?**

We hypothesize that there will be a significant difference in species
richness and structural diversity between habitats and vegetation types.
Previous research in Hawai’i Volcanoes National Park shows that there is
a sharp transition between wet forests and subalpine shrubland at 2000
millimeter mean annual precipitation. A Hawaiian treeline ecotones study
attributed the phenomenon to distinct changes in soil moisture
(Ainsworth and Drake 2024). We predict that both wet forest types will
have a more rapid increase in invasive species introduction over time
than Mauna Loa. This expectation is based on reduced physical barriers
(e.g., less steep terrain, increased connectivity with other habitat
types), higher resource availability, more human/animal activity to
disperse propagules, and a wider range of ecological niches for invasive
species to occupy.

# Dataset Information

Raw data was derived from the United States Government's open data site
starting from October 4th, 2010 to September 7th, 2023. This dataset is
part of The Pacific Island Inventory and Monitoring Network (PACN)
Established Invasive Plant Species (EIPS) Monitoring Protocol. More
information on the sources used can be found in the table below:

```{r dataset_information, fig.cap="Dataset Information on PACN EIPS Cover Class Data", echo=FALSE, message=FALSE, warning=FALSE}

#Create a dataframe with dataset information
Dataset.Information <- data.frame (
  Detail = c("Name", "Publisher", "Retrieved From", "Variables Used", "Date Range"),
  Description = c("Pacific Island Network Established Invasive Plant Species Monitoring Dataset 2010-2022", "National Park Service", "https://catalog.data.gov/dataset/pacific-island-network-established-invasive-plant-species-monitoring-dataset-2010-2022", "Plant Community, Sampling Frame, Decimal Latitude, Decimal Longitude, Date, Species Code, Scientific Name, Vernacular Name, Vegetation Type", "October 4, 2010 to September 7, 2023")
)

#Create a table 
knitr::kable(Dataset.Information,
             col.names = c("Detail", "Description"),
             digits = 2)
```

# Exploratory Analysis

## Exploration of Raw Data

The raw dataset included 37977 observations across 33 variables,
including sampling date, plant community (e.g., wet forest, subalpine
shrubland), transect number and type, transect latitude and longitude,
datum, species, and life form (which we later refer to as “vegetation
type”), among others. There were also several NA values which were
subsequently omitted from our processed dataset.

```{r RawData, echo=FALSE, message=FALSE, warning=FALSE}
CoverClass_Data <- read.csv(
  here::here("Data", "Raw", "PACN_EIPS_Cover_Class_Data.csv"), stringsAsFactors = FALSE)

str(CoverClass_Data)
```

## Study Site

Three regions within Hawai‘i Volcanoes National Park were selected as
the Big Island contained the most sampling sites out of all the Hawaiian
Islands in the dataset. All invasive plant species were counted within 5
× 20 m plots at each site, allowing for comparison between wet forest
and subalpine shrubland habitats. Figure \@ref(fig:Image) displays a
Landsat mosaic of the approximate location and average elevation of our
three study regions. In Figure \@ref(fig:MapofSamplingSites), sampling
sites are color-coded by region. Studies characterize wet forests with
greater wet canopy evaporation, annual mean precipitation and
temperature. Subalpine shrublands are found at higher elevations with
cooler temperatures and experience greater soil evaporation (Ainsworth
and Drake 2024). Mauna Loa, the subalpine shrubland, is depicted in
orange. Wet forest habitat types are shown in blue, with Nāhuku/East
Rift in light blue and Ola‘a in dark blue.

```{r Image, fig.cap="Landsat Mosaic of Big Island of Hawaii, Average Elevation (m)", out.width = "50%", echo=FALSE, message=FALSE, warning=FALSE}

knitr::include_graphics(
  here::here("Data", "Processed", "BigIsland.png")
)

```

```{r MapofSamplingSites, fig.cap="Map showing the sampling sites are color-coded by region", echo=FALSE, message=FALSE, warning=FALSE}

BigIsland_Processed <- 
  read.csv("Data/Processed/BigIsland_Processed.csv", stringsAsFactors = FALSE)
  
hawaii.counties <- st_read("Data/Raw/cb_2018_us_county_20m.shp")%>%
filter(STATEFP == 15 )

big_island <- hawaii.counties %>% 
  filter(NAME == "Hawaii")

sites_sf <- st_as_sf(BigIsland_Processed, coords = c("Decimal_Longitude","Decimal_Latitude"), crs = 4326) 

ggplot() +
  geom_sf(data=big_island, fill="lightgreen",
          color="forestgreen") +
  geom_sf(data = sites_sf, 
          aes(color = BigIsland_Processed$Sampling_Frame), 
          size = 2) +
  scale_color_manual(
  values = c(
    "Mauna Loa" = "orange2",
    "Nahuku / East Rift" = "lightblue",
    "Olaa" = "steelblue" ))+
  hawaiitheme+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title = "Sampling Locations on the Big Island",
       color = "Sampling Sites")
```

## Data Wrangling

Spatially, the raw dataset ranged from 19.4 to 22.0 degrees latitude and
-155.0 to -160.0 degrees longitude and was collected in WGS 1984 (4326).
To map our desired extent, we filtered the state of Hawai’i by counties
to extract just the Big Island, the result extent was between 19.0 to
20.2 degrees latitude and -154.8 to -156.0 degrees longitude.

After preliminary raw data exploration, we filtered for the three
sampling sites (Mauna Loa, Nahuku/East Rift, and Ola’a) on the Big
Island of Hawai’i. We selected the following variables that better suit
our analyses and renamed each column by removing “custom\_” prefix:
Plant Community, Sampling Frame, Decimal Latitude, Decimal Longitude,
Date, Species Code, Scientific Name, Vernacular Name, and Vegetation
Type. The wrangled, processed dataset was saved as
“BigIsland_Processed.csv”. Another dataset, “BigIsland_Clean.csv”, was
used to conduct ANOVA and Tukey Post-Hoc tests as these statistical
analyses required transect numbers.

# Analysis

```{r ResearchQ1, fig.cap="Overall Invasive Species Richness at Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

#Note: In the title names I added "\n" to create new line breaks to prevent text from disappearing from the screen, so they are NOT typos!
Invasive_Richness <- ggplot(BigIsland_Processed, aes(x = Sampling_Frame, fill = Sampling_Frame)) +
 geom_bar() +
  labs(
    title = "Overall Invasive Species Richness at Each Site\non the Big Island of Hawai'i",
    subtitle = "Wet Forest in Blue | Subalpine Shrubland in Orange",
    x = "Sample Region",
    y = "Species Count"
  ) +
  scale_fill_manual(values = c("Mauna Loa" = "orange2",
                    "Olaa" = "steelblue",
                    "Nahuku / East Rift" = "lightblue"),
                    guide = "none")
  
print(Invasive_Richness)
```

**Site Region Richness:** Our results support our hypothesis that Mauna
Loa had the lowest overall invasive species richness compared to other
regions. Higher elevation environments such as Mauna Loa tend to
experience less invasive species; likely due to harsher environmental
conditions, short growing seasons, and steep terrain that limit their
persistence. Notably, Ola’a has the lowest average elevation and the
greatest invasive species richness. Lower elevations tend to have milder
climates that support favorable conditions for diverse species growth
and proliferation (Ainsworth and Drake 2024).

Our ANOVA test revealed that there was a significant difference in
overall invasive species richness between Ola’a and Mauna Loa (p-value
of 0.028). There was no meaningful significance in invasive species
richness between Nahuku / East Rift and Mauna Loa (p-value of 0.161) and
Ola’a and Nahuku / East Rift (p-value of 0.803). We chose to define
richness as the number of distinct species types at a given transect by
using the function “n_distinct”, negating repeated counts of species
occurrence. In doing so, we were able to assign richness scores to each
transect and averaged across study regions to determine their respective
invasive species richness. After running a Tukey Post-Hoc Test, we found
that Ola’a has an overall higher invasive species richness relative to
Mauna Loa.

```{r ResearchQ2, fig.cap="Invasive Species Richness by Vegetation Type at Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Invasive_Richness_Type1 <- ggplot(BigIsland_Processed, aes(x = Vegetation_Type, fill = Sampling_Frame)) +
 geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Invasive Species Richness by Vegetation Type\nat Each Site on the Big Island of Hawai'i",
    subtitle = "Wet Forest in Blue | Subalpine Shrubland in Orange",
    x = "Vegetation Type",
    y = "Species Count"
    ) +
   scale_fill_manual(values = c("Mauna Loa" = "orange2",
                    "Olaa" = "steelblue",
                    "Nahuku / East Rift" = "lightblue"))
  
print(Invasive_Richness_Type1)
```

**Regional Structural Diversity:** We evaluated the differences in
vegetation types for non-native species richness for each site. Grass
and herbaceous species have the largest species count in all sites.
Figures \@ref(fig:fern), \@ref(fig:grass), \@ref(fig:herb),
\@ref(fig:rush), \@ref(fig:sedge), \@ref(fig:shrub), \@ref(fig:tree),
\@ref(fig:treefern), and \@ref(fig:vine) break down the species count by
each vegetation type with invasive species code abbreviations commonly
used at the Hawai’i Volcanoes National Park.

An ANOVA test analyzing the significance of structural differences
between study regions found that Nahuku / East Rift, Ola’a, and Mauna
Loa all had significant differences in structural diversity. By
aggregating the number of vegetation types found at each
transect/sampling point, we are able to determine the structural
diversity of a given point and translate regional patterns into
comparable data that is then analyzed for statistical significance (i.e.
a site with tree + shrub + fern + grass receives a diversity score of 4,
whereas a site with only grass + shrub receives diversity score of 2). A
Tukey Post-Hoc test revealed that Ola’a has a greater structural
diversity than Mauna Loa with a mean difference of 4.527. Nahuku-East
Rift is also more structurally diverse than Mauna Loa with a mean
difference of 3.081. Ola’a also has a higher structural diversity than
Nahuku / East Rift with a mean difference of 1.446, though this
difference is comparably less pronounced. This suggests that overall,
Ola’a has the greatest structural diversity, followed by Nahuku / East
Rift, and Mauna Loa with the lowest structural diversity.

## Time Series of Invasive Plants

```{r ResearchQ3, fig.cap="2010 to 2023 Detection of Invasive Species for Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Richness_by_Year <- BigIsland_Processed %>%
  mutate(Year = year(Date)) %>%
  group_by(Year, Sampling_Frame) %>%
  summarise(TotalSpecies = n_distinct(Species_Code), .groups = "drop")

#add linear interpolation
Richness_by_Year <- Richness_by_Year %>%
  arrange(Year) %>%
  group_by(Sampling_Frame) %>%
  mutate(TotalSpecies = zoo::na.approx(TotalSpecies, x = Year, na.rm = FALSE)) %>%
  ungroup()

invasion.time.plot <- ggplot (Richness_by_Year, aes(x = Year, y = TotalSpecies, color = Sampling_Frame)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  geom_smooth (method = "lm", se = FALSE, col = "coral3") +
  facet_wrap( ~ Sampling_Frame, ncol = 1) +
  scale_x_continuous(breaks = seq(2010, 2023, by = 2))+
  labs(
    title = "2010 to 2023 Detection of Invasive Species for Each Site\non the Big Island of Hawai'i",
    x = "Year",
    y = "Total Species Count",
    color = "Site"
  ) +
  scale_color_manual(values = c("Mauna Loa" = "orange2",
                    "Olaa" = "steelblue",
                    "Nahuku / East Rift" = "lightblue"))
invasion.time.plot
```

## Specific Invasive Species Map Exploration

```{r SpecificSpeciesMappingExploration, fig.cap="Locations of Highly Invasive Species on the Big Island", echo=FALSE, message=FALSE, warning=FALSE}

HighlyInvasive_subset <- 
  BigIsland_Processed %>%
  filter(Species_Code %in% c("JUNEFF","SENMAD","HEDGAR","TIBHER",
                             "RUBELL","MORFAY","PSICAT","SPHCOO","PASTAR"))

sites_sf <- st_as_sf(HighlyInvasive_subset, coords = c("Decimal_Longitude","Decimal_Latitude"), crs = 4326) 

ggplot() +
  geom_sf(data=big_island, fill="lightgreen",
          color="forestgreen") +
  geom_sf(data = sites_sf, 
          aes(color = HighlyInvasive_subset$Species_Code), 
          size = 2) +
  hawaiitheme+
   theme(plot.title = element_text(face = "bold",
                              size = 14,
                              color = "black",
                              family = "serif"))+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(title = "Locations of Highly Invasive Species on the Big Island",
       color = "Invasive Species Code")
```

## Diversity of invasive species across habitat type {.tabset}

### FERN

```{r fern, fig.cap="Invasive Fern Species Abundance at Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Fern_only <- BigIsland_Processed %>%
  filter(Vegetation_Type == "Fern")

Fern_Abundance <- ggplot(Fern_only, aes(x = Species_Code, fill = Species_Code)) +
  geom_bar() +
  facet_wrap(~ Sampling_Frame, ncol = 1) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Invasive Fern Species Abundance at Each Site on the\nBig Island of Hawai'i",
    x = "Fern Species Type",
    y = "Species Count"
  ) +
  theme(legend.position = "none")

Fern_Abundance
```

### GRASS

```{r grass, fig.cap="Invasive Grass Species Abundance at Each Site on the\nBig Island of Hawai'i", echo = FALSE, message = FALSE, warning = FALSE}

Grass_only <- BigIsland_Processed %>%
  filter(Vegetation_Type == "Grass")

Grass_Abundance <- ggplot(Grass_only, aes(x = Species_Code, fill = Species_Code)) +
  geom_bar() +
  facet_wrap(~ Sampling_Frame, ncol = 1) +
  labs(
    title = "Invasive Grass Species Abundance at Each Site on the\nBig Island of Hawai'i",
    x = "Grass Species Type",
    y = "Species Count"
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 270))

Grass_Abundance
```

### HERB

```{r herb, fig.cap="Invasive Herb Species Abundance at Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Herb_only <- BigIsland_Processed %>%
  filter(Vegetation_Type == "Herb")

Herb_Abundance <- ggplot(Herb_only, aes(x = Species_Code, fill = Species_Code)) +
  geom_bar() +
  facet_wrap(~ Sampling_Frame, ncol = 1) +
  labs(
    title = "Invasive Herb Species Abundance at Each Site on the\nBig Island of Hawai'i",
    x = "Herb Species Type",
    y = "Species Count"
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 270))

Herb_Abundance
```

### RUSH

```{r rush, fig.cap="Invasive Rush Species Abundance at Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Rush_only <- BigIsland_Processed %>%
  filter(Vegetation_Type == "Rush")

Rush_Abundance <- ggplot(Rush_only, aes(x = Species_Code, fill = Species_Code)) +
  geom_bar() +
  facet_wrap(~ Sampling_Frame, ncol = 1) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Invasive Rush Species Abundance at Each Site on the\nBig Island of Hawai'i",
    x = "Rush Species Type",
    y = "Species Count"
  ) +
  theme(legend.position = "none")

Rush_Abundance
```

### SEDGE

```{r sedge, fig.cap="Invasive Sedge Species Abundance at Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Sedge_only <- BigIsland_Processed %>%
  filter(Vegetation_Type == "Sedge")

Sedge_Abundance <- ggplot(Sedge_only, aes(x = Species_Code, fill = Species_Code)) +
  geom_bar() +
  facet_wrap(~ Sampling_Frame, ncol = 1) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Invasive Sedge Species Abundance at Each Site on the\nBig Island of Hawai'i",
    x = "Sedge Species Type",
    y = "Species Count"
  ) +
  theme(legend.position = "none")

Sedge_Abundance
```

### SHRUB

```{r shrub, fig.cap="Invasive Shrub Species Abundance at Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Shrub_only <- BigIsland_Processed %>%
  filter(Vegetation_Type == "Shrub")

Shrub_Abundance <- ggplot(Shrub_only, aes(x = Species_Code, fill = Species_Code)) +
  geom_bar() +
  facet_wrap(~ Sampling_Frame, ncol = 1) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Invasive Shrub Species Abundance at Each Site on the\nBig Island of Hawai'i",
    x = "Shrub Species Type",
    y = "Species Count"
  ) +
  theme(legend.position = "none")

Shrub_Abundance
```

### TREE

```{r tree, fig.cap="Invasive Tree Species Abundance at Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Tree_only <- BigIsland_Processed %>%
  filter(Vegetation_Type == "Tree")

Tree_Abundance <- ggplot(Tree_only, aes(x = Species_Code, fill = Species_Code)) +
  geom_bar() +
  facet_wrap(~ Sampling_Frame, ncol = 1) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Invasive Tree Species Abundance at Each Site on the\nBig Island of Hawai'i",
    x = "Tree Species Type",
    y = "Species Count"
  )+
  theme(legend.position = "none")

Tree_Abundance
```

### TREE FERN

```{r treefern, fig.cap="Invasive Tree Fern Species Abundance at Each Site on the Big Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Tree.Fern_only <- BigIsland_Processed %>%
  filter(Vegetation_Type == "Tree Fern")

Tree.Fern_Abundance <- ggplot(Tree.Fern_only, aes(x = Species_Code, fill = Species_Code)) +
  geom_bar() +
  facet_wrap(~ Sampling_Frame, ncol = 1) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Invasive Tree Fern Species Abundance at Each Site on the\nBig Island of Hawai'i",
    x = "Tree Fern Species Type",
    y = "Species Count"
  ) +
  theme(legend.position = "none")

Tree.Fern_Abundance
```

### VINE

```{r vine, fig.cap="Invasive Vine Species Abundance at Each Site on the\nBig Island of Hawai'i", echo=FALSE, message=FALSE, warning=FALSE}

Vine_only <- BigIsland_Processed %>%
  filter(Vegetation_Type == "Vine")

Vine_Abundance <- ggplot(Vine_only, aes(x = Species_Code, fill = Species_Code)) +
  geom_bar() +
  facet_wrap(~ Sampling_Frame, ncol = 1) +
  scale_fill_brewer(palette = "Paired") +
  labs(
    title = "Invasive Vine Species Abundance at Each Site on the\nBig Island of Hawai'i",
    x = "Vine Species Type",
    y = "Species Count"
  ) +
  theme(legend.position = "none")

Vine_Abundance
```

Figure \@ref(fig:ResearchQ3) displays a total invasive species count
over time across study regions. This count includes multiple
observations of the same species at a given sample site. It appears that
while levels have fluctuated in the past 12-13 years, there has been an
overall decrease in invasive species counts across all three regions.
Mauna Loa is experiencing the greatest decrease with about 3 less
invasive species counted annually. Nahuku / East Rift is declining by
approximately 0.6558 species per year. Ola’a has experienced little to
no fluctuation in the number of invasive species with an average decline
of 0.1059 species annually. One important feature to note are the points
on the graph representing the times that each region was sampled. Given
that these sampling windows are limited and occurred sporadically
throughout the time span, trends observed are likely not very
substantial but do warrant further investigation.

# Summary and Conclusions

Our results show that invasive species richness does change across the
sample regions. Figure \@ref(fig:ResearchQ1) displays the overall
invasive species richness across the 13-year sampling protocol. Ola’a
appears to have the greatest species richness across the study regions,
followed by Nahuku / East Rift. Mauna Loa has the least species richness
out of the three regions. This finding was somewhat supported by our
ANOVA (analysis of variance) and Tukey Post-Hoc test, which found that
Ola’a has significantly higher invasive species richness in comparison
to Mauna Loa. There is no statistically significant difference between
Ola’a & Nahuku / East Rift and Mauna Loa and Nahuku / East Rift.

Our findings also suggest that structural diversity changes across all
three of the sample regions. Figure \@ref(fig:ResearchQ2) demonstrates
that Mauna Loa only hosted invasive grasses, herbs, rushes, and shrubs
whereas Nahuku / East Rift and Ola’a supported some combination of all
nine of possible vegetation types. Our analysis via Tukey multiple
comparisons of means on an ANOVA test reveals that there are
statistically significant differences between all three regions. Mauna
Loa has the least structural diversity of the three regions, followed by
Nahuku / East Rift with an intermediary level of structural diversity.
Ola’a is the most diverse in number of vegetation types at a given
sampling point.

These phenomena are likely attributed to:

-   Higher resource availability in wet forest habitat types

-   Less human/animal activity to disperse propagules in subalpine
    shrubland habitats

-   Wider range of ecological niches for invasive species to occupy in
    wet forest habitat types

-   Altitude, soil moisture, and wind pattern differences among habitat
    types

-   Reduced physical barriers in wet forest habitat types

    -   Less steep terrain

    -   Increased connectivity with other habitat types

However, further analysis and data regarding elevation, rainfall, wind,
and other potential drivers are required to confirm these theories as
they pertain to our specific dataset. While we had two wet forest
habitats to compare, our analysis would be made stronger if we had
additional subalpine shrubland regions to confirm if the patterns we
observed on Mauna Loa can be extrapolated to this habitat type as a
whole on the Big Island of Hawai’i.

Trends varied across time. Throughout the three study regions, the
number of invasive plants counted decreased on average. Though our time
series results are likely anecdotal, some claims are somewhat
substantiated by research conducted on the Big Island. One paper found
that “while most of the area occupied by invasive species increased in
size, the area of two highly invasive species occupying upper elevation
wet forest decreased dramatically” (Cordell 2021). A useful next step in
exploring these trends would be to have a longer timeline of invasive
species spread. While our findings suggest that invasive species
populations are largely stable across the study regions, this may be an
outlier in the long-term trends of the island. Further, as we only have
three study regions to pull data from, we are only able to speak to the
local trends at these sites. Perhaps the increased protectivity within
the Hawai’i Volcanoes National Park has allowed these regions to
maintain or even decrease invasive species counts, whereas the rest of
the Big Island has an entirely different experience.

Understanding regional trends in invasive species populations is key to
informing best practices for land management, tailoring restoration
techniques on a case-by-case basis. Invasive species aren’t the only
threat to the endemic plants of Hawai’i, instead “extinction is
projected for some native plant populations when habitat area is defined
by climatic envelopes is impinged by sea level rise on the lower end and
elevation at the upper end” (Fortini et al., 2013). This may put
high-elevation sites such as Mauna Loa at greater risk of losing native
plants, which is only exacerbated by added competition from invasive
species.

# References

Ainsworth, A., & Drake, D. R. (2024). Hawaiian treeline ecotones:
Implications for plant community conservation under climate change.
Plants, 13(1), 123.

Cordell, S. (2020). Hawaii and US-affiliated Pacific Islands: Regional
summaries. In S. J. Seybold, S. J. Frankel, D. H. Olson, R. Flitcroft,
B. K. Kerns, J. Lee, K. L. Ripley, & A. S. Munson (Eds.), Invasive
species in forests and grasslands of the United States: A comprehensive
ecological assessment. U.S. Department of Agriculture, Forest Service.
<https://www.fs.usda.gov/research/publications/book/invasiveSpecies/invasiveSpeciesAppendixHawaiiPacificIslands.pdf>

Ewel, J. J., O'Dowd, D. J., Bergelson, J., Daehler, C. C., D'Antonio, C.
M., Gómez, L. D., Gordon, D. R., Hobbs, R. J., Holt, A., & Hopper, K. R.
(1999). Deliberate introductions of species: Research needs, benefits
can be reaped, but risks are high. BioScience, 49(8), 619–630.

Fortini, L., Price, J., Jacobi, J., Vorsino, A., Burgett, J., Brinck,
K., Amidon, F., Miller, S., Gon III, S. ʻO., Koob, G., & Paxton, E.
(2013, November). A landscape-based assessment of climate change
vulnerability for all native Hawaiian plants (Technical Report
HCSU-044). Hawai‘i Cooperative Studies Unit, University of Hawai‘i at
Hilo.
<https://hilo.hawaii.edu/hcsu/documents/TR44_Fortini_plant_vulnerability_assessment.pdf>

Hawaiʻi Department of Agriculture & Biosecurity. (2011). Final EA posted
for biological control of invasive strawberry guava [Press release].
Hawaiʻi Department of Agriculture.
<https://dab.hawaii.gov/blog/news-releases/final-ea-posted-for-biological-control-of-invasive-strawberry-guava/>

Hawaiʻi Invasive Species Council. (2023). Invasive species profiles
[Website]. Hawaiʻi Department of Land and Natural Resources.
<https://dlnr.hawaii.gov/hisc/info/invasive-species-profiles/>

National Park Service. (2021, February 18). Ferns – Hawaiʻi Volcanoes
National Park [Web page]. U.S. Department of the Interior.
<https://www.nps.gov/havo/learn/nature/ferns.htm>

National Park Service, Pacific Island Inventory & Monitoring Network.
(2025, November 25). Pacific Island Network established invasive plant
species monitoring dataset, 2010–2022 [Data set]. Data.gov.
<https://catalog.data.gov/dataset/pacific-island-network-established-invasive-plant-species-monitoring-dataset-2010-2022>

Young, H. S., McCauley, D. J., & Dunne, J. A. (2016). The role of
species interactions in food web stability: A multi-level approach.
Canadian Journal of Zoology, 94(12), 939–949.

```{r add report to our repository, include = FALSE}
#COPY/PASTE IN CONSOLE to update and get html in our working directory files
#rmarkdown::render("MilesflynnRoybal_ENV872_EDA_FinalProject.Rmd", output_format = "bookdown::html_document2")
```
