---
title: "Comparison of Invasive Plant Species and Topography in the Big Island, Hawaii"
author: "Environmental Data Analytics | Elizabeth Miles-Flynn & Irina Roybal"
subtitle: "Mauna Loa, Olaa, and Nahuku/East Rift"
date: "December 10, 2025"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    lof: true
    lot: true
    theme: cerulean
    highlight: tango
---

```{r setup, include = FALSE}
#Setup
library(tidyverse)
library(lubridate)
library(here)
library(knitr)

#Check Working Directory
getwd()
here()
```

### 0.1 Git Repository

<https://github.com/irinaroybal/MilesflynnRoybal_ENV872_EDA_FinalProject.git>

## 1 Rationale and Research Questions

The purpose of this study is to analyze the spatial relationship of invasive species across three different study sites on the Big Island of Hawaii. Before Polynesian contact with the remote Pacific, invasive species colonized islands at a rate of one species every 50,000 years. Once Polynesians arrived, this rate increased to three to four species every hundred years (Young 2016). Since European colonization, this rate has erupted to around twenty invasive species introduced each year in Hawaii (Ewell 1999). Invasive species, defined as a non-native organism that spreads rapidly and/or causes harm to the environment, are a considerable threat to the native biota of Hawaii. Of the Hawaiian Islands’ 10,000 native species, nearly 90% are endemic (Cordell 2021). This includes about 1,367 endemic taxa of vascular plants. 

Our goal is to evaluate the present-day rate of invasive species introduction in the Big Island over the past decade to inform best management practices to protect endemic species of the Hawaiian Islands. Our dataset includes three sites on the Big Island: Mauna Loa, Ola’a, and Nahuku. Mauna Loa is classified as a subalpine shrubland, and Ola’a and Nahuku are considered wet forests. Our dataset ranges from 2010 to 2023 and contains coordinates for each site. Through our analysis, we aim to answer the following research questions:

1.  Does diversity of invasive species change across different habitat types (wet forest and subalpine shrubland)?

2.  How does invasive species concentration and diversity change over time?

## 2 Dataset Information
Raw data was derived from the United States Government's open data site starting from October 4th, 2010 to September 7th, 2023. This dataset is part of The Pacific Island Inventory and Monitoring Network (PACN) Established Invasive Plant Species (EIPS) Monitoring Protocol. More information on the sources used can be found in the table below: 

```{r dataset.information, fig.cap = "Table 2.1: Dataset Information on PACN EIPS Cover Class Data", echo = FALSE}

#Create a dataframe with dataset information
Dataset.Information <- data.frame (
  Detail = c("Name", "Publisher", "Retrieved From", "Variables Used", "Date Range"),
  Description = c("Pacific Island Network Established Invasive Plant Species Monitoring Dataset 2010-2022", "National Park Service", "https://catalog.data.gov/dataset/pacific-island-network-established-invasive-plant-species-monitoring-dataset-2010-2022", "Plant Community, Sampling Frame, Decimal Latitude, Decimal Longitude, Date, Species Code, Scientific Name, Vernacular Name, Life Form", "October 4, 2010 to September 7, 2023")
)

#Create a table 
knitr::kable(Dataset.Information,
             col.names = c("Detail", "Description"),
             caption = "Table 2.1: Dataset Information on PACN EIPS Cover Class Data",
             digits = 2)
```
## 3 Exploratory Analysis



### 3.1 Exploration of Raw Data
The raw dataset included 33 total variables focus on describing the status of non-native plant richness, frequency, and cover. 
```{r Raw Data, echo = FALSE}
CoverClass_Data <- 
  read.csv("Data/Raw/PACN_EIPS_Cover_Class_Data.csv", stringsAsFactors = FALSE)

str(CoverClass_Data)
```
### 3.2 Study Site
Out of five Hawaiian national parks, we only selected three sites with wet forest and subalpine shrublands habitats that followed the same methodology of counting all non-native plant species found within 5-by-20 meter plots at each site.

### 3.3 Data Wrangling

After preliminary raw data exploration, we wrangled and saved the processed dataset as “BigIsland.Rmd” to better suit our subsequent analyses. 

## 4 Analysis

### 4.1 Diversity of invasive species across habitat type

### 4.2 Time series of invasive plants 


## 5 Summary and Conclusions

## 6 References
