---
title: "Comparison of Invasive Plant Species and Topography on the Big Island of Hawai'i"
author: "Environmental Data Analytics | Elizabeth Miles-Flynn & Irina Roybal"
subtitle: "Mauna Loa, Ola'a, and Nahuku/East Rift"
date: "December 10, 2025"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    lof: true
    lot: true
    theme: cerulean
    highlight: tango
---

```{r setup, include = FALSE}
#Setup
library(tidyverse)
library(lubridate)
library(here)
library(knitr)

#Check Working Directory
getwd()
here()
```

### 0.1 Git Repository

<https://github.com/irinaroybal/MilesflynnRoybal_ENV872_EDA_FinalProject.git>

## 1 Rationale and Research Questions

The purpose of this study is to analyze the spatial distribution of invasive species across three study sites on the Big Island of Hawai‘i. Prior to Polynesian arrival in the remote Pacific, new invasive species established naturally at an estimated rate of one species every 50,000 years. After Polynesian settlement, this rate increased to roughly three to four species every hundred years (Young 2016). Following European contact, the introduction rate accelerated dramatically to approximately twenty invasive species per year in Hawai‘i (Ewell 1999).

While non-native species include any organisms introduced outside their natural range, only a subset of them become invasive. Invasive species are defined as non-native organisms that spread rapidly and/or cause ecological, economic, or cultural harm. These invasive species pose a substantial threat to Hawai‘i’s unique native biota. Of the roughly 10,000 native species found across the Hawaiian Islands, nearly 90% are endemic—occurring nowhere else on Earth. 

Our goal is to evaluate the invasive species on the Big Island across three study sites over the past thirteen years to inform best management practices to protect endemic species of the Hawaiian Islands. Our dataset focuses on three sites on the Big Island: Mauna Loa, Ola’a, and Nahuku/East Rift. Mauna Loa is classified as a subalpine shrubland, and Ola’a and Nahuku/East Rift are considered wet forest habitats. Our dataset ranges from 2010 to 2023 and contains coordinates for each survey site where a transect was used to count species-level observations and counts of repeated occurrences. Through our analysis, we aim to answer the following research questions:


1. Does invasive species' richness and abundance change across the three sample regions?

2. How does the frequency and distribution of invasive species change over time for each site?


We hypothesize that there will be a significant difference in species richness and abundance between wet forest habitats and the subalpine shrubland habitat. Previous research in Hawai’i Volcanoes National Park shows that there is a sharp transition between wet forests and subalpine habitats. A Hawaiian treeline ecotones study attributed the phenomenon to distinct changes in altitude, soil moisture, and wind patterns (Ainsworth and Drake 2023). We predict that both wet forest types will have a more rapid increase in invasive species introduction over time than Mauna Loa. This expectation is based on reduced physical barriers (e.g., less steep terrain, increased connectivity with other habitat types), higher resource availability, more human/animal activity to disperse propagules, and a wider range of ecological niches for invasive species to occupy.

## 2 Dataset Information
Raw data was derived from the United States Government's open data site starting from October 4th, 2010 to September 7th, 2023. This dataset is part of The Pacific Island Inventory and Monitoring Network (PACN) Established Invasive Plant Species (EIPS) Monitoring Protocol. More information on the sources used can be found in the table below: 

```{r dataset.information, fig.cap = "Table 2.1: Dataset Information on PACN EIPS Cover Class Data", echo = FALSE}

#Create a dataframe with dataset information
Dataset.Information <- data.frame (
  Detail = c("Name", "Publisher", "Retrieved From", "Variables Used", "Date Range"),
  Description = c("Pacific Island Network Established Invasive Plant Species Monitoring Dataset 2010-2022", "National Park Service", "https://catalog.data.gov/dataset/pacific-island-network-established-invasive-plant-species-monitoring-dataset-2010-2022", "Plant Community, Sampling Frame, Decimal Latitude, Decimal Longitude, Date, Species Code, Scientific Name, Vernacular Name, Life Form", "October 4, 2010 to September 7, 2023")
)

#Create a table 
knitr::kable(Dataset.Information,
             col.names = c("Detail", "Description"),
             caption = "Table 2.1: Dataset Information on PACN EIPS Cover Class Data",
             digits = 2)
```
## 3 Exploratory Analysis



### 3.1 Exploration of Raw Data
The raw dataset included 33 total variables focus on describing the status of non-native plant richness, frequency, and cover. 
```{r Raw Data, echo = FALSE}
CoverClass_Data <- 
  read.csv("Data/Raw/PACN_EIPS_Cover_Class_Data.csv", stringsAsFactors = FALSE)

str(CoverClass_Data)
```
### 3.2 Study Site
Out of five Hawaiian national parks, we only selected three sites with wet forest and subalpine shrublands habitats that followed the same methodology of counting all non-native plant species found within 5-by-20 meter plots at each site.

### 3.3 Data Wrangling

After preliminary raw data exploration, we wrangled and saved the processed dataset as “BigIsland.Rmd” to better suit our subsequent analyses. 

## 4 Analysis

### 4.1 Diversity of invasive species across habitat type

### 4.2 Time series of invasive plants 


## 5 Summary and Conclusions

## 6 References
